Remark on the hypothetical judgment

### 2015-08-01 paper.tex
fix typos; clarify ontological status of verification objects

In 1956, Heyting propounded his version of what has come to be known as the
\emph{Brouwer-Heyting-Kolmogorov} interpretation of intuitionistic logic, by
explaining the assertion conditions of the propositions. Note that where Heyting
says ``assert a proposition'', in light of Martin-L\"of's clarification, we must
understand ``assert \emph{the truth of} a proposition''. Heyting's explanation
of the assertion conditions for the truth of implication were as follows:
\begin{quote}
  The implication $\Imp{p}{q}$ may be asserted if and only if we possess a
  construction $r$, which, joined to any construction proving $p$ (supposing the
  latter be effected), would automatically effect a construction of $q$. \cite{HeytingA:int}
\end{quote}

@@

This, however, may be too much to ask, since in modern proof-theoretic accounts
of meaning, the hypothetical judgment as explained by Martin-L\"of must be
understood in a very strong sense, where its proof shall be an object with a
``hole'' in it, which could be plugged with a proof for the antecedent to yield
a proof for the consequent: that is, it is not enough that one should have a way
of transforming the proof of the antecedent into proof for the consequent, but
one must have a uniform way to do so.

@@

If we return to Heyting's original definition of the assertion
$\IsTrue{\Imp{P}{Q}}$, it is clear that he would have to accept any mathematical
means of transforming the construction of $\IsTrue{P}$ into a construction of
$\IsTrue{Q}$, since the construction $r$ in his definition is not constrained by
any uniformity condition: it is merely any effective operation which, when
adjoined with a construction of the premise, effects a construction of the
conclusion; and moreover, $r$ may proceed by virtue of any of the transcendental
observations which are possible by virtue of the truth of $P$ having been
experienced by the subject.

And what of Martin-L\"of? In his type theory, only the introduction rules for
the types are given, and the elimination rules are ``theorems'' (or admissible
rules) which are evident only in the non-uniform sense; that is, the
justifications of the elimination rules proceed by introspection on the possible
ways in which their premises were experienced. Therefore, if Martin-L\"of's
uniform hypothetical judgment is to be accepted, it is necessary that the
statement of the elimination rules be effected using something that permits
a non-uniform demonstration.

@@

Why? This purports to be a proof of the judgment
$\HypJ{\IsTrue{P}}{\IsTrue{P\land Q}}$, which is not evident under the uniform
explanation of hypothetical judgment. It could be made evident if the definition
of a proposition consisted in the declaration of both its introduction rules and
its elimination rules, but in fact, only the introduction rules are given, and
the elimination rules are simply codifications of common patterns of inference.

Indeed, Martin-L\"of justifies the elimination rules using the non-uniform
notion of consequence. This rule, for instance, is justified as follows:

@@

If we are to take the explanation of the elimination rules seriously, then, we
must read Martin-L\"of as having already at his disposal a kind of hypothetical
judgment whose evidence consists in any effective means at all of transforming
the evidence for the premise into the evidence for the conclusion (we might call
this ``material consequence'' as opposed to ``logical consequence''), so the use
of the uniform hypothetical judgment elsewhere has not somehow removed the need
for this kind of inference.

@@

In fact, by replacing the abstract/synthetic explanations of the propositions in
terms of their verification acts, we can recast them as giving rise to a theory
of sets or types by explaining them in terms of their verification objects (i.e.
their canonical witnesses).

@@

\begin{gather*}
  \infer={
    \EqSet{A}{B}
  }{
    \Eval{A}{A'} &
    \Eval{B}{B'} &
    \InputMode{\VAL{A'}}\equiv\InputMode{\VAL{B'}}
  }\qquad
  \infer={
    \IsSet{A}
  }{
   \EqSet{A}{A}
  }\\
  \infer={
    \EqMember{M}{N}{A}
  }{
    \Eval{A}{A'} &
    \Eval{M}{M'} &
    \Eval{N}{N'} &
    \VAL{A'}(\InputMode{M}, \InputMode{N})
  }\qquad
  \infer={
    \Member{M}{A}
  }{
    \EqMember{M}{M}{A}
  }
\end{gather*}

>>>>

In 1956, Heyting propounded his version of what has come to be known as the
\emph{Brouwer-Heyting-Kolmogorov} interpretation of intuitionistic logic, by
explaining the assertion conditions of the propositions. Note that where Heyting
says ``assert a proposition'', in light of Martin-L\"of's clarification, we must
understand ``assert \emph{the truth of} a proposition''. Heyting's explanation
of the assertion conditions for the truth of implication were as follows:
\begin{quote}
  The implication $\Imp{\mathfrak{p}}{\mathfrak{q}}$ may be asserted if and only
if we possess a construction $\mathfrak{r}$, which, joined to any construction
proving $\mathfrak{p}$ (supposing the latter be effected), would automatically
effect a construction of $\mathfrak{q}$. \cite{HeytingA:int}
\end{quote}

@@

This, however, may be too much to ask, since in modern proof-theoretic accounts
of meaning, the hypothetical judgment as explained by Martin-L\"of must be
understood in a very strong sense, where its proof shall be an object with a
``hole'' in it, which could be plugged with a proof for the antecedent to yield
a proof for the consequent: that is, it is not enough that one should have a way
of transforming the proof of the antecedent into a proof of the consequent, but
one must have a uniform way to do so.

@@

If we return to Heyting's original definition of the assertion
$\IsTrue{\Imp{P}{Q}}$, it is clear that he would have to accept any mathematical
means of transforming the construction of $\IsTrue{P}$ into a construction of
$\IsTrue{Q}$, since the construction $\mathfrak{r}$ in his definition is not
constrained by any uniformity condition: it is merely any effective operation
which, when adjoined with a construction of the premise, effects a construction
of the conclusion; and moreover, $\mathfrak{r}$ may proceed by virtue of any of
the transcendental observations which are possible by virtue of the truth of $P$
having been experienced by the subject.

And what of Martin-L\"of? In his type theory, only the introduction rules for
the types are given, and the elimination rules are ``theorems'' (or admissible
rules) which are evident only in the non-uniform sense; that is, the
justifications of the elimination rules proceed by introspection on the possible
ways in which their premises could have been experienced. Therefore, if
Martin-L\"of's uniform hypothetical judgment is to be accepted, it is necessary
that the statement of the elimination rules be effected using something that
permits a non-uniform demonstration.

@@

Why? This purports to be a proof of the judgment
$\HypJ{\IsTrue{P}}{\IsTrue{P\land Q}}$, which is not evident under the uniform
explanation of hypothetical judgment. It could be made evident if the definition
of a proposition consisted in the declaration of both its introduction rules and
its elimination rules, but in fact, only the introduction rules are given, and
the elimination rules are simply codifications of common patterns of reasoning
from premise to conclusion.

Indeed, Martin-L\"of justifies the elimination rules using the non-uniform
(material, rather than logical) consequence. This rule, for instance, is
justified as follows:

@@

If we are to take the explanation of the elimination rules seriously, then, we
must read Martin-L\"of as having already at his disposal a kind of hypothetical
judgment whose evidence consists in any effective means at all of transforming
the demonstration of the premise into a demonstration of the conclusion, so the
use of the uniform hypothetical judgment elsewhere has not somehow saved us the
need to explain this kind of inference.

@@

In fact, we can replace the abstract/synthetic explanations of the propositions
in terms of their verification acts with new explanations in terms of
verification objects (i.e. their canonical witnesses); then, the verification
act consists in constructing a verification object.

A proposition is verified just in case there exists a verification object, but
it is important to understand that this is not to say that a (possibly unknown)
verification object may exist outside the subject's experience (construction) of
it. Rather, this is a trivial equivalence, since to say that an object exists it
the same as to say that the subject has constructed it.\footnote{The idea that
verification or proof objects exist separately from our experience of them is
part of the realist ontology which is now espoused by Martin-L\"of
\cite{prawitz:2012}, contrary to his position at the time of the Siena lectures;
this view of course cannot be accepted by Brouwerians, who profess a thoroughly
idealist ontology \cite{sundholm-mva, sundholm:2014}.}

@@

\begin{gather*}
  \infer={
    \EqSet{A}{B}
  }{
    \Eval{A}{A'} &
    \Eval{B}{B'} &
    \InputMode{\VAL{A'}}\equiv\InputMode{\VAL{B'}}
  }\qquad
  \infer={
    \IsSet{A}
  }{
   \EqSet{A}{A}
  }\\
  \infer={
    \EqMember{M}{N}{A}
  }{
    \Eval{A}{A'} &
    \Eval{M}{M'} &
    \Eval{N}{N'} &
    \VAL{A'}(\InputMode{M'}, \InputMode{N'})
  }\qquad
  \infer={
    \Member{M}{A}
  }{
    \EqMember{M}{M}{A}
  }
\end{gather*}
